---
title: "Horse Racing Data Analysis Project - ML1 HS24"
author: "sindhu.chinnamsetti@stud.hslu.ch, ronald.schwalb@stud.hslu.ch, eniko.simo@stud.hslu.ch"
date: "Jan 5, 2025"
output:
  html_document:
    toc: True
    toc_depth: 2
    df_print: paged
  word_document: default
  pdf_document: default
editor_options: 
  markdown: 
    wrap: sentence
---

# 1) Data Cleaning
<span style="color:red">WIP</span>


```{r}
# Import the Data from CSV Files
df1 <- read.csv("Lingfield_AW_2014_2024_flat.csv")
df2 <- read.csv("Lingfield_AW_2014_2024_jumps.csv")
df3 <- read.csv("Lingfield_turf_2014_2024_flat.csv")
df4 <- read.csv("Lingfield_turf_2014_2024_jumps.csv")

df.horse <- rbind(df1, df2, df3, df4)

str(df.horse)
head(df.horse)
```

- gotta diminish dataset size to: 
- max 10^5 = 100,000 rows
- 10-20 predictors

Keep the following columns based on domain knowledge and careful inspection of the dataset:

```{r}
df.horse <- subset(df.horse, select=c(date, race_name,type, class, dist_m, going, ran, pos, draw, horse, age, sex, lbs, hg, secs, jockey, trainer, prize))

str(df.horse)
head(df.horse)
```

Create a new column 'won', assigning 1 if 'pos' is 1, otherwise 0

```{r}
df.horse$won <- ifelse(df.horse$pos == 1, 1, 0)
# Check the first few rows to ensure the new column has been added
head(df.horse[c("pos","won")],50)
# => looks good
```

Check for missing values. NAs and empty strings
```{r}
colSums(is.na(df.horse) | df.horse == "")
```

```{r}
# let's drop hg and prize because most of them are missing
# but let's keep draw because AFAIK it might influence outcome
# keep it in mind that it only exists for Flat course type

df.horse <- subset(df.horse, select=-c(hg,prize))
```
Check datatypes. 
```{r}
str(df.horse)
```
Some variable types should be modified for better analysis. 

Convert 'date' column to date format
```{r}
df.horse$date <- as.Date(as.character(df.horse$date), format = "%Y-%m-%d")
```


```{r}
# check which values of pos can't be converted to integers
unique(df.horse$pos)
```

Meaning of the abbreviations:

**UR** (Unseated Rider): The jockey fell off during the race.<br>
**PU** (Pulled Up): The horse was stopped by the jockey during the race due to an issue, usually to prevent injury.<br>
**F** (Fell): The horse fell during the race.<br>
**DSQ** (Disqualified): The horse was disqualified, usually for breaking rules like interference.<br>
**RR** (Refused to Race): The horse refused to start the race.<br>
**BD** (Brought Down): The horse was knocked down by another horse, usually in a fall.<br>
**RO** (Ran Out): The horse ran off the course, failing to follow the designated track.<br>
**SU** (Slipped Up): The horse slipped and fell during the race, usually on a turn or due to ground conditions.<br>
**REF** (Refused): The horse refused to jump a fence or hurdle, or it refused to continue racing at some point.<br>
Each of these indicates why a horse didn't complete the race as expected. Thus, these rows can be removed. 

Remove rows where pos is not a number:
```{r}
df.horse <- df.horse[!df.horse$pos %in% c("UR", "PU", "F", "DSQ", "RR", "BD", "RO", "SU", "REF"), ]
```

See if it worked:

```{r}
unique(df.horse$pos)
```
 

```{r}
# check the same for secs
df.horse$secs_num <- as.numeric(df.horse$secs)
df.horse[is.na(as.integer(df.horse$secs_num)), ]$secs
```
```{r}
# see rows where secs = "-"
df.horse[df.horse$secs=="-",]
```
This looks like a legit entry but since we have no time data for it, let's remove it. 

```{r}
df.horse <- df.horse[df.horse$secs != "-", ]
```

Now convert them to numeric variables.

```{r}
df.horse$pos <- as.integer(df.horse$pos)
df.horse$secs <- as.numeric(df.horse$secs)
```

Convert all characters to factors:
```{r}
# Convert all character columns to factors
df.horse[sapply(df.horse, is.character)] <- lapply(df.horse[sapply(df.horse, is.character)], as.factor)

str(df.horse)
```
```{r}
# remove helper variable, won't need it anymore
df.horse <- subset(df.horse, select=-secs_num)
# check if it worked
colnames(df.horse)
```

- SHOULD "WON" BE TURNED INTO A BOOLEAN? 
- SHOULD "POS" BE AN ORDERED FACTOR? (ChatGPT says no, but prof said we should have all var types...?)


Log-transform “amounts” -> at prof's recommendation
In our dataset, these variables can be considered amounts:
- dist_m
- lbs
- secs

HOW ABOUT POS ???

```{r}
# Log-transforming 'dist_m', 'lbs', and 'secs'
df.horse$log_dist_m <- log(df.horse$dist_m)
df.horse$log_lbs <- log(df.horse$lbs)
df.horse$log_secs <- log(df.horse$secs)
```

Visualize original variable + log to see if distribution is more balanced.
```{r}
par(mfrow = c(3, 2))

hist(df.horse$dist_m)
hist(df.horse$log_dist_m)
hist(df.horse$lbs)
hist(df.horse$log_lbs)
hist(df.horse$secs)
hist(df.horse$log_secs)

```
<br>
TODO: MAKE THIS PRETTIER AND DELETE THE BELOW

```{r}
hist(df.horse$dist_m)
```

```{r}
hist(df.horse$log_dist_m)
```
<br>
LOOKS WEIRD ...

```{r}
hist(df.horse$lbs)
```
```{r}
hist(df.horse$log_lbs)
```
```{r}
hist(df.horse$secs)
```
```{r}
hist(df.horse$log_secs)
```
<br>
I DON'T KNOW... LOGS DON'T LOOK A WHOLE LOT MORE BALANCED THAN ORIGINAL VARIABLES. 
WHAT SHOULD I SEE HERE?? 


## Sample

Running the models take forever with our current dataset, so create a randomized sample with 10% of the data so that we can see results faster at this early stage of the project. 

TODO: USE STRATIFIED SAMPLING!

```{r}
# Sample 10% of the dataset for testing
#df.horse_sample <- df.horse[sample(nrow(df.horse), 0.1 * nrow(df.horse)), ]

#dim(df.horse_sample)
```
OR:

Running the models take forever with our current dataset, so create a stratified sample with 10% of the data so that we can see results faster at this early stage of the project. 


```{r}
# Install the rsample package if needed
#install.packages("rsample")

# Load required libraries
library(rsample)

# Assume df.horse is your data frame, and you're stratifying by 'type'
set.seed(123)  # Set seed for reproducibility

# Create a 10% stratified sample based on the 'type' column
split <- initial_split(df.horse, prop = 0.1, strata = type)

# Get the training set (this will be your 10% stratified sample)
stratified_sample <- training(split)

# Check the distribution to ensure stratification worked
table(stratified_sample$type)  # This should be proportional to the original distribution

# View the sampled data
head(stratified_sample)
```
```{r}
df.horse_sample <-stratified_sample
dim(df.horse_sample)
```
See if the sample is proportionate

```{r}
print("df.horse$type:")
summary(df.horse$type)
print("df.horse_sample$type:")
summary(df.horse_sample$type)
```
LOOKS PRETTY GOOD. IS THERE A BETTER WAY TO CHECK THIS?
MAYBE VISUALIZE? 



# 2) Linear Model
WIP
```{r}
plot(log_secs ~ log_lbs, data = df.horse_sample,
     main = "Log secs against log weight")
```

Would expect a positive correlation: the more weight a horse carries 
-> the longer it will take to complete the race. 
Somewhat visible. 

```{r}
plot(log_dist_m ~ log_lbs, data = df.horse_sample,
     main = "Log distance against log weight")
```

```{r}
plot(log_secs ~ log_dist_m, data = df.horse_sample,
     main = "Log time against log distance")
```

The longer the distance the longer time it takes. 
Logically sound and clearly visible. 

```{r}
plot(log_secs ~ sex, data = df.horse_sample,
     main = "Log time against sex")
```

HAHA, WHY ARE THERE 6? 
TODO: Add explanation

```{r}
plot(log_dist_m ~ sex, data = df.horse_sample,
     main = "Log distance against sex")
```
```{r}
plot(log_secs ~ won, data = df.horse_sample,
     main = "Log time against winning")
```

THIS IS NOT GOOD <br>
A BETTER CHART IS NEEDED? MAYBE SWAP X + Y AXES?
OR KEEP THIS TO A MODEL WHERE Y IS BINARY.



# 3) Generalised Linear Model - Poisson
Placeholder

# 4) Generalised Linear Model - Binomial
WIP 
```{r}
glm.horse_sample <- glm(won ~ type + log_dist_m + going + age + sex + log_lbs + log_secs, 
                data = df.horse_sample,
                family = "binomial")
summary(glm.horse_sample)

```
<span style="color:red";>HOW TO INTERPRET ALL THESE FACTOR VARIABLES?
SHOULD THEY ALL BE FACTORS OR IS THERE A BETTER WAY?
I DON'T REMEMBER SO MANY FACTORS FROM THE LECTURES. </span>



# 5) Generalised Additive Model
Placeholder

# 6) Neural Network
Placeholder

# 7) Support Vector Machine
Placeholder

# 8) Use of Generative AI
Placeholder

# 9) Conclusion
Placeholder