---
title: "Horse Racing Data Analysis Project - ML1 HS24"
author: "sindhu.chinnamsetti@stud.hslu.ch, ronald.schwalb@stud.hslu.ch, eniko.simo@stud.hslu.ch"
date: "Jan 5, 2025"
output:
  html_document:
    toc: True
    toc_depth: 2
    df_print: paged
  word_document: default
  pdf_document: default
editor_options: 
  markdown: 
    wrap: sentence
---

# 1) Data Cleaning
<span style="color:red">WIP</span>


```{r}
# Import the scraped racingpost Data from CSV Files
df1 <- read.csv("Lingfield_AW_2014_2024_flat.csv")
df2 <- read.csv("Lingfield_AW_2014_2024_jumps.csv")
df3 <- read.csv("Lingfield_turf_2014_2024_flat.csv")
df4 <- read.csv("Lingfield_turf_2014_2024_jumps.csv")

df.horse <- rbind(df1, df2, df3, df4)

str(df.horse)
head(df.horse)
```


```{r}
# Import the Betfair Starting Price data
df.bsp <- read.csv("ML1_BSP_DF.csv")
str(df.bsp)
head(df.bsp)

```
```{r}
# Merge df.horse and df.bsp

# Split the 'horse' column into 'horse' and 'origin'
df.horse <- df.horse %>%
  mutate(
    origin = gsub(".*\\((.*)\\)", "\\1", horse),  # Extract the text inside parentheses
    horse = gsub("\\s*\\(.*\\)", "", horse)      # Remove the text inside parentheses and the brackets
  )

# In both the racingost and BSP dataset, the names are converted to upper cases to have a higher chance of matching when joining.
df.horse$horse <-
  toupper(gsub("[' ]", "", df.horse$horse))
df.bsp$HORSE_NAME <-
  toupper(gsub("[' ]", "", df.bsp$HORSE_NAME))

# Now, merge the two data frames on the keys 'horse'and 'date'
df.horses <-
  merge(df.horse,
        df.bsp,
        by.x = c("horse", "date"),
        by.y = c("HORSE_NAME", "DATE")
        )

# Find rows in df.bsp that did not match with df.horse
unmatched_horses_bsp <- anti_join(df.bsp, df.horse, 
                                  by = c("HORSE_NAME" = "horse", "DATE" = "date"))

# A total of 215 horses did not merge due to slight mismatches in the spelling of horse names.
# We've noticed here that in the racingpost scraped data for some reason an I was added at the end of the horse's name e.g. SNOWKING became SNOWKINGI
# Utilizing Chatgpt, this could have been rectified without plenty manual fixing, see below:

# Get and print unique horse names
unique_horses <- unmatched_horses_bsp %>%
  distinct(HORSE_NAME) %>%
  pull(HORSE_NAME)

# Update the horse names
df.horse <- df.horse %>%
  mutate(horse = ifelse(horse %in% paste0(unique_horses, "I"), 
                        substr(horse, 1, nchar(horse) - 1), 
                        horse))

# Now, merge the two data frames on the keys 'horse'and 'date'
df.horses <-
  merge(df.horse,
        df.bsp,
        by.x = c("horse", "date"),
        by.y = c("HORSE_NAME", "DATE")
        )

```
- There were 150 instances where datasets could not be merged due to mismatches in horse's
names
- Now all rows are removed where the counted number of horses for the event is less than indicated by column HORSE_COUNT

```{r}
# Count occurrences of each unique date and off combination
df.horses <- df.horses %>%
  group_by(date, off) %>%                 # Group by date and off
  mutate(horse_count2 = n()) %>%          # Count occurrences within each group
  ungroup()                               # Remove grouping

# Remove rows where horse_count2 < HORSE_COUNT
df.horses <- df.horses %>%
  filter(horse_count2 >= HORSE_COUNT)


```

- gotta diminish dataset size to: 
- max 10^5 = 100,000 rows
- 10-20 predictors
Keep the following columns based on domain knowledge and careful inspection of the dataset:

```{r}
df.horse <- subset(df.horses, select=c(horse, date, course, off, type, class, dist_m, going, ran, pos, draw, age, sex, lbs, secs, dec, jockey, trainer, or, rpr, sire, dam, damsire, owner, origin, WIN, BSP_WIN, Imp_Perc_WIN, PLACE, BSP_PLACE, Imp_Perc_PLACE))

str(df.horse)
head(df.horse)
```

### Dataset Column Descriptions

c(PLACE, BSP_PLACE, Imp_Perc_PLACE)

-   **horse**: The name of the horse.
-   **date**: The date on which the race occurred.
-   **course**: The track on which the race took place (AW stands for All Weather which refers to the track type)
-   **off**: The time the race was set to begin.
-   **type**: The race type e.g. flat, hurdles or chase.
-   **class**: The class of the horse.
-   **dist_m**: The race distance in meters.
-   **going**: The state of the track or racecourse at the time of the event.
-   **ran**: The total number of participants in the race.
-   **pos**: The finishing position of the horse in the race.
-   **draw**: The starting position or gate assigned to a horse at the beginning of the race.
-   **age**: The age of the horse participating in the race.
-   **sex**: The gender of the horse.
-   **lbs**: The amount of weight in pounds the horse carries during the race, including any penalties or allowances.
-   **secs**: The time in seconds to complete the race.
-   **dec**: The decimal odds associated with a horse's chances of winning.
-   **jockey**: The name of the jockey riding the horse during the race.
-   **trainer**: The name of the trainer responsible for preparing the horse for the race.
-   **or**: The official rating of a horse, indicating its ability based on past performances.
-   **rpr**: The Racing Post Rating, a numerical assessment of a horse's performance.
-   **sire**: The sire or father of the horse, providing pedigree information.
-   **dam**: The dam or mother of the horse, offering additional pedigree details.
-   **damsire**: The grandfather of the horse, further contributing to its lineage.
-   **owner**: The name of the person or organisation owning the horse.
-   **origin**: The country where the horse was bred.
-   **WIN**: The result indicating whether the horse won the race.
-   **BSP_WIN**: The Betfair Starting Price or Odds for the horse to win at the time the race begins (inverse of BSP_WIN).
-   **Imp_Perc_WIN**: The implied probability percentage based on the BSP for the selected horse.
-   **PLACE**: The result indicating whether the horse placed in the top positions.
-   **BSP_PLACE**: The Betfair Starting Price or Odds for the horse placing at the time the race begins.
-   **Imp_Perc_PLACE**: The implied probability percentage for the horse placing at the time the race begins (inverse of BSP_PLACE).

Check for missing values. NAs and empty strings
```{r}
colSums(is.na(df.horse) | df.horse == "")
```


Check datatypes. 
```{r}
str(df.horse)
```
Some variable types should be modified for better analysis. 

Convert 'date' column to date format
```{r}
df.horse$date <- as.Date(as.character(df.horse$date), format = "%Y-%m-%d")
```


```{r}
# check which values of pos can't be converted to integers
unique(df.horse$pos)
```

Meaning of the abbreviations:

**UR** (Unseated Rider): The jockey fell off during the race.<br>
**PU** (Pulled Up): The horse was stopped by the jockey during the race due to an issue, usually to prevent injury.<br>
**F** (Fell): The horse fell during the race.<br>
**DSQ** (Disqualified): The horse was disqualified, usually for breaking rules like interference.<br>
**RR** (Refused to Race): The horse refused to start the race.<br>
**BD** (Brought Down): The horse was knocked down by another horse, usually in a fall.<br>
**RO** (Ran Out): The horse ran off the course, failing to follow the designated track.<br>
**SU** (Slipped Up): The horse slipped and fell during the race, usually on a turn or due to ground conditions.<br>
**REF** (Refused): The horse refused to jump a fence or hurdle, or it refused to continue racing at some point.<br>
Each of these indicates why a horse didn't complete the race as expected. Thus, these rows can be removed. 

Change rows where pos is not a number:
```{r}
# Abbreviation for non-finishers
values_to_replace <- c("UR", "PU", "F", "DSQ", "RR", "BD", "RO", "SU", "REF")

# Group by type, count specific values, and calculate total races
result <- df.horse %>%
  group_by(type) %>%
  summarise(
    count = sum(pos %in% values_to_replace),  # Count occurrences of specific values
    total_races = n()                      # Count total races for each type
  )

# View the result
print(result)

# Add the `finished` column and replace values in `pos`
df.horse <- df.horse %>%
  mutate(
    # Add a column indicating if the horse finished
    finished = ifelse(grepl("^[0-9]+$", pos), 1, 0),  # Check if `pos` is numeric
    # Replace specific values in `pos` with the corresponding value from `ran`
    pos = ifelse(pos %in% values_to_replace, ran, pos)
  )
```

See if it worked:

```{r}
unique(df.horse$pos)
```
 

```{r}
# check the same for secs
df.horse$secs_num <- as.numeric(df.horse$secs)
df.horse[is.na(as.integer(df.horse$secs_num)), ]$secs
```
```{r}
# see rows where secs = "-"
df.horse[df.horse$secs=="-",]
```
This looks like a legit entry but since we have no time data for it, let's remove it. 

```{r}
df.horse <- df.horse[df.horse$secs != "-", ]
```

Now convert them to numeric variables.

```{r}
df.horse$pos <- as.integer(df.horse$pos)
df.horse$secs <- as.numeric(df.horse$secs)
```

Convert all characters to factors:
```{r}
# Convert all character columns to factors
df.horse[sapply(df.horse, is.character)] <- lapply(df.horse[sapply(df.horse, is.character)], as.factor)

str(df.horse)
```
```{r}
# remove helper variable, won't need it anymore
df.horse <- subset(df.horse, select=-secs_num)
# check if it worked
colnames(df.horse)
```
Sex is a factor with 6 levels.
```{r}
unique(df.horse$sex)
```
Explanations:
Race horse sexes: 
**G (Gelding)**: A male horse that has been castrated. Geldings are often calmer and easier to handle, making them popular in racing.

**M (Mare)**: An adult female horse, usually aged 5 years or older (though this can vary by region).

**F (Filly)**: A young female horse, generally under 4 or 5 years old, depending on the racing jurisdiction.

**C (Colt)**: A young male horse, typically under 4 or 5 years old, that has not been castrated.

**H (Horse)**: This might specifically refer to a mature male horse that has not been castrated, often referred to as a "stallion" but used differently in racing contexts.

**R (Ridgling or Rig)**: A male horse with undescended testicles, meaning it is partially castrated or not fully developed sexually.

Create also a binary variable for sex. Let's keep both variables because the 6 different types might make a difference later.  


```{r}
# Rename the original varialbe
df.horse$sex_6 <- df.horse$sex
# Drop old sex column
df.horse <- subset(df.horse, select=-sex)
# Create binary sex variable
df.horse$sex <- ifelse(df.horse$sex_6 %in% c("G", "C", "H", "R"), "M", "F")
# Convert to factor
df.horse$sex <- as.factor(df.horse$sex)
```

- SHOULD "WON" BE TURNED INTO A BOOLEAN? 
- SHOULD "POS" BE AN ORDERED FACTOR? (ChatGPT says no, but prof said we should have all var types...?)


Log-transform “amounts” -> at prof's recommendation
In our dataset, these variables can be considered amounts:
- dist_m
- lbs
- secs
- age

HOW ABOUT POS ???

```{r}
# Log-transforming 'dist_m', 'lbs', and 'secs'
df.horse$log_dist_m <- log(df.horse$dist_m)
df.horse$log_lbs <- log(df.horse$lbs)
df.horse$log_secs <- log(df.horse$secs)
df.horse$log_age <- log(df.horse$age)
```

Visualize original variable + log to see if distribution is more balanced.
```{r}
par(mfrow = c(4, 2))

hist(df.horse$dist_m)
hist(df.horse$log_dist_m)
hist(df.horse$lbs)
hist(df.horse$log_lbs)
hist(df.horse$secs)
hist(df.horse$log_secs)
hist(df.horse$age)
hist(df.horse$log_age)

```

<br>
I DON'T KNOW... LOGS DON'T LOOK A WHOLE LOT MORE BALANCED THAN ORIGINAL VARIABLES. 
WHAT SHOULD I SEE HERE?? 
SHOULDN'T THEY BE CENTERED AROUND 0??

```{r}
str(df.horse)
```

## Sample

Running the models take forever with our current dataset, so create a randomized sample with 10% of the data so that we can see results faster at this early stage of the project. 

TODO: USE STRATIFIED SAMPLING!

```{r}
# Sample 10% of the dataset for testing
#df.horse_sample <- df.horse[sample(nrow(df.horse), 0.1 * nrow(df.horse)), ]

#dim(df.horse_sample)
```
OR:

Running the models take forever with our current dataset, so create a stratified sample with 10% of the data so that we can see results faster at this early stage of the project. 


```{r}
# Install the rsample package if needed
#install.packages("rsample")

# Load required libraries
library(rsample)

# Assume df.horse is your data frame, and you're stratifying by 'type'
set.seed(123)  # Set seed for reproducibility

# Create a 10% stratified sample based on the 'type' column
split <- initial_split(df.horse, prop = 0.1, strata = type)

# Get the training set (this will be your 10% stratified sample)
stratified_sample <- training(split)

# Check the distribution to ensure stratification worked
table(stratified_sample$type)  # This should be proportional to the original distribution

# View the sampled data
head(stratified_sample)
```
```{r}
df.horse_sample <-stratified_sample
dim(df.horse_sample)
```
See if the sample is proportionate

```{r}
print("df.horse$type:")
summary(df.horse$type)
print("df.horse_sample$type:")
summary(df.horse_sample$type)
```
LOOKS PRETTY GOOD. IS THERE A BETTER WAY TO CHECK THIS?
MAYBE VISUALIZE? 



# 2) Linear Model
WIP
```{r}
plot(log_secs ~ log_lbs, data = df.horse_sample,
     main = "Log secs against log weight")
```

Would expect a positive correlation: the more weight a horse carries 
-> the longer it will take to complete the race. 
Somewhat visible. 

```{r}
plot(log_dist_m ~ log_lbs, data = df.horse_sample,
     main = "Log distance against log weight")
```

```{r}
plot(log_secs ~ log_dist_m, data = df.horse_sample,
     main = "Log time against log distance")
```

The longer the distance the longer time it takes. 
Logically sound and clearly visible. 

```{r}
plot(log_secs ~ sex, data = df.horse_sample,
     main = "Log time against sex")
```

 

```{r}
plot(log_dist_m ~ sex, data = df.horse_sample,
     main = "Log distance against sex")
```

```{r}
plot(log_lbs ~ sex, data = df.horse_sample,
     main = "Log weight against sex")
```

THESE LAST 3 DON'T BELONG TO LINEAR MODELS!
BUT LAB1 ALSO INCLUDES THESE...

Now let's consider 2 predictors in a single graph. 
```{r}
plot(log_dist_m ~ log_secs, data = df.horse_sample,
     col=sex,
     pch=19,
     main = "Log distance against log time")
legend("topleft",
       pch=19,
       legend=c("M","F"), # THESE ARE MANUALLY PROVIDED -> HOW CAN I KNOW WHICH IS WHICH???
       col=c("blue","red"))
```

```{r}
library(ggplot2)
qplot(y= log_dist_m,x=log_secs,
data=df.horse_sample,
facets= ~sex)
```

```{r}
plot(log_dist_m ~ log_secs, data = df.horse_sample,
     col=sex,
     pch=19,
     main = "Log distance against log time")
```
```{r}
boxplot(log_dist_m ~ won, data = df.horse_sample,
     main = "Log distance against won")
```
```{r}
boxplot(log_secs ~ won, data = df.horse_sample,
     main = "Log time against won")
```

Plotting against the "won" variable doesn't really show anything. 

NEXT STEP: fit models
(see LinearModels1:CatsLab - 3 Fitting models)



# 3) Generalised Linear Model - Poisson
Placeholder

# 4) Generalised Linear Model - Binomial
WIP 
```{r}
glm.horse_sample <- glm(won ~ type + log_dist_m + going + age + sex + log_lbs + log_secs, 
                data = df.horse_sample,
                family = "binomial")
summary(glm.horse_sample)

```
HOW TO INTERPRET ALL THESE FACTOR VARIABLES?
SHOULD THEY ALL BE FACTORS OR IS THERE A BETTER WAY?
I DON'T REMEMBER SO MANY FACTORS FROM THE LECTURES.



# 5) Generalised Additive Model
Placeholder

# 6) Neural Network
Placeholder

# 7) Support Vector Machine
WIP

Trying to find a good categorical target variable. 
Check the Distribution of sex_6: 

```{r}
table(df.horse$sex_6)
```
The distribution of the sexes looks rather imbalanced. 
Since the rare categories (H, M, R) don’t have meaningful distinctions, combine them into an "Other" category:
```{r}
df.horse$sex_4 <- as.character(df.horse$sex_6)  # Convert to character to modify levels
df.horse$sex_4[df.horse$sex_4 %in% c("H", "M", "R")] <- "Other"  # Combine rare categories
df.horse$sex_4 <- as.factor(df.horse$sex_4)  # Convert back to factor

# Check the new distribution
table(df.horse$sex_4)
```
#TODO:
#CONTINUE WITH THIS:
Stratified Sampling for Training
Ensure that your training data is representative by using stratified sampling. This prevents the model from being biased toward majority classes during training:

```{r}
#library(caret)

## Stratified train-test split
#trainIndex <- createDataPartition(df$sex_6, p = 0.7, list = FALSE)
#df_train <- df[trainIndex, ]
#df_test <- df[-trainIndex, ]
```


# 8) Use of Generative AI
Placeholder

# 9) Conclusion
WIP

- Conclusions were a lot less clear to draw from this real-world dataset than in the case of the sample datasets presented at the lectures.